\pkg{regsem}: Regularized Structural Equation Modeling

### 2017-03-04 draft1.tex
formatted with rmd

---
author:
  - name: Ross Jacobucci
    affiliation: University of Notre Dame
    address: >
      First line
      Second line
    email: rcjacobuc@gmail.com
    url: rjacobucci.com
title:
  formatted: "\\pkg{regsem}: Regularized Structural Equation Modeling"
  # If you use tex in the formatted title, also supply version without
  plain:     "regsem: Regularized Structural Equation Modeling"
  # For running headers, if needed
  short:     "\\pkg{regsem}"
abstract: >
  The \pkg{regsem} package in \proglang{R}, an implementation of Regularized Structural Equation Modeling [@jacobucci2016regularized], was recently developed with the goal incorporating various forms of penalized likelihood estimation with a broad array of structural equations models. The forms of regularization include both the \textit{ridge} [@hoerl1970] and the least absolute shrinkage and selection operator (\textit{lasso}; [@Tibshirani1996], along with sparser extensions. The paper details both the algorithmic details and an overview of the use of \pkg{regsem} through both a factor analysis and latent growth curve models. 
keywords:
  # at least one keyword must be supplied
  formatted: [regularization, structural equation modeling, latent variables, "\\proglang{R}"]
  plain:     [regularization, structural equation modeling, latent variables, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{algorithm}
  \usepackage[noend]{algpseudocode}
  \usepackage[subnum]{cases}

output: rticles::jss_article
bibliography: jss_regsem.bib
---
\setlength\parindent{24pt}
\setlength{\parskip}{0.5em}

# Introduction

The desire for simplicity in model structure comes by many names, including simple structure [@thurstone1937], variable complexity [@browne2001], parsimony [@raykov1999desirability; @marsh1996assessing], "sparse loadings" in the context of principal components analysis [@zou2006sparse], and lastly, "sparsistency", denoting that all parameters in a sparse model that are zero are correctly estimated as zero with probability tending to one [@lam2009sparsistency]. The goal is accurately and efficiently estimate a model that is parsimonious in allowing users to easily interpret the model's representation of reality. In Figure 1, this would entail testing whether the cross-loadings are necessary to the fit of the model, or whether they can be pared away without an meaningful increase in bias.

@@

```{r, eval=FALSE, include=FALSE}
library(lavaan)
library(semPlot)
library(semTools)

sim.mod <- "
i =~ 1*x1 + 1*x2 + 1*x3 + 1*x4
s =~ 0*x1 + 1*x2 + 2*x3 + 3*x4
i ~ 1*c1 + .2*c2 + 0*c3 + 0*c4 + 0*c5 + 0*c6 + 0*c7 + 0*c8+ 0*c9+ 0*c10
s ~ 1*c1 + .2*c2 + 0*c3 + 0*c4 + 0*c5 + 0*c6 + 0*c7 + 0*c8+ 0*c9+ 0*c10"

dat <- simulateData(sim.mod,model.type="growth",sample.nobs=80,seed=1234)

mod1 <- "
i =~ 1*x1 + 1*x2 + 1*x3 + 1*x4
s =~ x1 + 1*x2 + 2*x3 + 3*x4
i ~ c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8+ c9+ c10
s ~ c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8+ c9+ c10
"

out <- growth(mod1,dat,fixed.x=T)
summary(out)
fitmeasures(out)

semPaths(out, whatLabels="est",intercepts=FALSE,residuals=TRUE,exoCov=FALSE)
```

@@

As an example, a researcher may want to test this model, but may only have a relatively small sample size (e.g. 80). However, there are 29 estimated parameters in this model, resulting in a estimated parameter to sample size ratio as far below even the most liberal recommendations [e.g. 10:1 parameters to sample size; @kline2015principles]. In lieu of finding additional respondents, reducing the number of parameters estimated is one effective strategy. Specifically, the 20 estimated regressions from \textit{c1-c10} could be reduced to a number that makes the number of parameters estimated to sample size ratio more reasonable.

# Regularization

Although a whole host of methods exist to perform variable selection, the use of regularization has seen a wide array of application in the context of regression, and more recently, in areas such as graphical modeling, as well as a host of others. The two most common procedures for regularization in regression are the \textit{ridge} [@hoerl1970] and the least absolute shrinkage and selection operator [\textit{lasso}; @Tibshirani1996]; however, there are various alternative forms that can be seen as subsets or generalizations of these two procedures. Given an outcome vector \textit{y} and predictor matrix $X \in {R}^{n \times p}$ , the ridge estimates are defined as

$$\tag{1}
\hat{\beta}^{ridge}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p} \beta_{j}^{2}\Big\},
$$
\noindent
where $\beta_{0}$ is the intercept, $\beta_{j}$ is the coefficient for $x_{j}$, and $\lambda$ is the penalty that controls the amount of shrinkage.  Note that when $\lambda = 0$, equation 3 reduces to ordinary least squares regression. As $\lambda$ is increased, the $\beta$ parameters are shrunken towards zero. The lasso estimates are defined as

$$\tag{2}
\hat{\beta}^{lasso}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p}|\beta_{j}|\Big\}.
$$

\noindent
In lasso regression, the $l_{1}$-norm is used, instead of $l_{2}$-norm as in ridge, which also shrinks the $\beta$ parameters, but additionally drives the parameters all the way to zero, thus performing a form of subset selection.

In the context of our example depicted in Figure 1, to use lasso regression to select among the covariates, the growth model would need to be reduced to two factor scores, which neglects both the relationship between both the slope and intercept, reducing both to independent factor scores. Particularly in models with a greater number of latent variables, this becomes increasing problematic. A method that keeps the model structure, while allowing for a penalized estimation of specific parameters is regularized structural equation modeling [RegSEM; @jacobucci2016regularized]. RegSEM adds a penalty function to the traditional maximum likelihood cost function for SEMs. The maximum likelihood cost function for SEMs can be written as

$$\tag{3}
F_{ML}=log(\left|\Sigma\right|)+tr(C*\Sigma^{-1})-log(\left|C\right|)- p.
$$

\noindent
where $\Sigma$ is the model implied covariance matrix, $C$ is the observed covariance matrix, and $p$ is the number of estimated parameters. RegSem builds in an additional element to penalize certain model parameters yielding

$$\tag{4}
F_{regsem} = F_{ML} + \lambda P(\cdot)
$$

\noindent
where $\lambda$ is the regularization parameter and takes on a value between zero and infinity. When $\lambda$ is zero, MLE is performed, and when $\lambda$ is infinity, all penalized parameters are shrunk to zero. $P(\cdot)$ is a general function for summing the values of one or more of the model's parameter matrices. The two common forms of $P(\cdot)$ include both the Lasso ($\| \cdot \|_{1}$), which penalizes the sum of the absolute values of the parameters, and Ridge ($\| \cdot \|_{2}$), which penalizes the sum of the squared values of the parameters.

In our example, the twenty regression parameters from the covariates to both the intercept and slope would be penalized. Using lasso penalties, the absolute value of these twenty parameters would be summed and after being multiplied by the penalty $\lambda$, added to equation 4, resulting in:

$$\tag{5}
F_{lasso} = F_{ML} + \lambda * \left\| 	\begin{matrix}	
c1 \xrightarrow[]{} i\\
c2\xrightarrow[]{}i\\
\vdots \\
c10\xrightarrow[]{}i\\
c1\xrightarrow[]{}s\\
\vdots \\
c10\xrightarrow[]{}s\\
\end{matrix}  \right\|_{1}
$$

@@

## Optimization

One method that has become popular for optimizing penalized likelihood method is that of proximal gradient descent [e.g. p. 104 in @hastie2015statistical]. In comparison to one-step procedures common in SEM optimization, that only involve a method for calculating the step size and the direction (typically using the gradient and an approximation of the Hessian), proximal gradient descent can be formulated as a two-step procedure. With a stepsize of $\alpha^{t}$ and parameters $\theta^{t}$ at iteration \textit{t}:

@@

where $S_{\alpha \lambda}(z)$ is the soft-thresholding operator [@donoho1995noising] used to overcome non-differentiability of the lasso penalty at the origin: 
\begin{equation}
S_{\alpha \lambda}(z_{j}) = sign(z_{j})(|z_{j}|-\alpha \lambda)_{+}
\end{equation}
where $(x)_{+}$ is shorthand for max(x,0) and $\alpha$ is the step size. Henceforth, $\lambda$ is assumed to encompass both the penalty and the step size $s^{t}$. This procedure is only used to updated parameters that are subject to penalty. Non-penalized parameters are updated only using step 1 from above.

@@

\noindent\begin{minipage}{\textwidth}
	%	\renewcommand\footnoterule{}    
	\begin{algorithm}[H]
		\begin{algorithmic}[1]
			%	\Procedure{RegSEM Block Coordinate Descent for Lasso}{}
			%	\footnotetext{Note: $A(pen)$ refers only to the penalized parameters, $vec()$ concatenates both vectors}
			\State Generate starting values for $\theta_{t}$
			\State Calculate initial fit $F_{t}$
			\State Set step size $\alpha$. .5 works well at this time.
			\State Set tolerance (tol). e.g. 1e-6
			\While {$|F_{t} - F_{t+1}| > tol$}
			\State Calculate gradient for A:$ \nabla (A) =:  \frac{\partial A}{\partial \theta_{t}}$
			\State $ \theta_{t+1^{*},A} =: \theta_{t,A} - \alpha \nabla(A)$
			\State  Update penalized parameters: $\theta_{t+1^{*},A(pen)} =: S_{\alpha \lambda}(\theta_{t+1^{*},A(pen)})$
			\State Calculate gradient for S:$ \nabla (S) =:  \frac{\partial S}{\partial \theta_{t+1^{*}}}$ 
			\State Update S parameters: $ \theta_{t+1^{*},S} =: \theta_{t,S} - \alpha \nabla(S)$
			\State $\theta_{t+1} =: \theta_{t+1^{*}}$ %vec(\theta_{t+1^{*},A},\theta_{t+1^{*},S})$
			\State Update $S_{t+1}, A_{t+1}$
			\State $ \Sigma_{t+1} = F(I-A_{t+1})^{-1}S_{t+1}(I-A_{t+1})^{-T}F^{T}$
			\State $ F_{t+1} = F_{ML}(\Sigma_{t+1},C) + \lambda \| A_{t+1}(pen) \|$
			\EndWhile
			%	\EndProcedure
		\end{algorithmic}
		\caption{RegSEM Block Coordinate Descent}
		\label{alg:seq}
	\end{algorithm}
\end{minipage}
\noindent
where step 13 is the calculation of the implied covariance matrix using RAM matrices and step 14 calculates the fit of the model with penalizing parameters in the \textit{A} matrix. Note that parameters from the \textit{S} matrix can also be penalized, however, this is much less common. This algorithm can be described as first order proximal block coordinate descent. For some SEM models, using block updates has been found to work better than standard gradient descent with the same soft thresholding of penalized parameters.

# Types of Penalties

@@

## Elastic Net

Most notably, the elastic net [@zou2005regularization] encompasses both the ridge and lasso, reaching a compromise between both through the addition of an additional parameter $\alpha$.

$$
P_{enet}(\theta_{j}) = 0.5(1-\alpha)\| \theta_{j} \|_{2} + \alpha\| \theta_{j} \|_{2}
$$

with a soft-thresholding update of 
$$
S(\theta_{j})= 
\begin{cases}
0,&  |\theta_{j}| < \alpha\lambda\\
\frac{sgn(\theta_{j})(|\theta_{j}|-\alpha\lambda)}{1+(1-\alpha)\lambda},              & |\theta_{j}|\geq\alpha\lambda
\end{cases}
$$

\noindent
When $\alpha$ is zero, ridge is performed, and conversely when $\alpha$ is 1, lasso regularization is performed. This method harnesses the benefits of both methods, particularly when variable selection is warranted (lasso), but there may be collinearity between the variables (ridge).

## Adaptive Lasso

In using lasso penalties, difficulties emerge when the scale of variables differ dramatically. By only using one value of $\lambda$, this can add appreciable bias to the resulting estimates [e.g. @fan2001variable]. One method proposed for overcoming this limitation is the adaptive lasso [@zou2006adaptive]. Instead of penalizing parameters directly, each parameter is scaled by the un-penalized estimated (maximum likelihood estimate in SEM). This manifests itself as:
$$
F_{alasso} = F_{ML} + \lambda \| \theta_{ML}^{-1} * \theta_{pen} \|_{1}
$$
\noindent
with, following the same form for the lasso, the soft-thresholding update is:
[\\]: ftp://ftp.stat.math.ethz.ch/Teaching/buhlmann/advanced-comput-statist/notes1.pdf
$$
S(\theta_{j})= sign(\theta_{j})(|\theta_{j}|-\frac{\lambda}{2|\theta_{j}|})_{+}
$$
\noindent
In this, larger penalties are given for non-significant (smaller) parameters, limiting the bias in estimating larger, significant, parameters. Note that one limitation of this approach for SEM models is that the model needs to be estimable with maximum likelihood. Particularly for models with large numbers of variables, in relation to sample size, this may not be possible.

## Smoothly Clipped Absolute Deviation Penalty
[\\]: High dimensional statistics p. 32

@@

Two additional penalties that overcome some of the deficiencies of the lasso, producing sparser solutions, include the smoothly clipped absolute deviation penalty [SCAD; @fan2001variable] and the minimax concave penalty [MCP; @zhang2010nearly]. In comparison to the lasso, both the SCAD and MCP have much steeper penalties for smaller parameters, as is evident in Figure **.

@@

$$
pen_{\lambda,\gamma}(\theta_{j}) = \lambda \big\{I(\theta_{j}\leq\lambda) + \frac{(\gamma \lambda-0)_{+}}{(\gamma-1)\lambda}I(\theta_{j}>\lambda)\big\}
$$

\noindent
with a soft-thresholding update of
$$
S(\theta_{j})= 
\begin{cases}
S(\theta_{j},\lambda),&  |\theta_{j}| \geq 2\lambda\\
\frac{\gamma-1}{\gamma-2}S(\theta_{j},\frac{\lambda\gamma}{\gamma-1}),              & 2\lambda < |\theta_{j}|\leq\alpha\lambda\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
$$

@@

## Minimax Concave Penalty

@zhang2010nearly proposed the minimax concave penalty:

$$
pen_{\lambda,\gamma}(\theta_{j}) = \lambda\bigg(|\theta_{j}|-\frac{\theta_{j}^{2}}{2\lambda\gamma}\bigg)I(|\theta_{j}|<\lambda\gamma) +\frac{\lambda^{2}\gamma}{2}I(|\theta_{j}|\geq \lambda\gamma)
$$

\noindent
with a soft-thresholding update of
$$
S(\theta_{j})= 
\begin{cases}
\frac{\gamma}{\gamma-1}S(\theta_{j},\lambda),&  |\theta_{j}| \leq \lambda\gamma\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
$$

@@

# Implementation

RegSEM is implemented as the \pkg{regsem} package in the \proglang{R} statistical environment. To estimate the maximum likelihood fit of the model, \pkg{regsem} uses \textit{Reticular Action Model} [RAM; @McArdle_1984, @mcardle2005] notation to derive an implied covariance matrix. The parameters of each SEM are translated into three matrices: the \textit{filter} (\textit{F}), the \textit{asymmetric} (\textit{A}; directed paths; e.g. factor loadings or regressions), and the \textit{symmetric} (\textit{S}; undirected paths; e.g. covariances or variances). See @jacobucci2016regularized for more detail on RAM notation.

Syntax for using the \pkg{regsem} is based on the \pkg{lavaan} package [@rosseel2012] for structural equation models. \pkg{lavaan} is a general SEM software program that can fit a wide array of models with various estimation methods. To use \pkg{regsem}, the user has to first fit the model in \pkg{lavaan}. As an example, below is the code for a one factor model fit in lavaan with the Holzinger-Swineford [@holzinger1939study] dataset.

```{r,message=FALSE}
library(lavaan)
mod <- "
f1 = ~ NA*x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
f1~~1*f1
"
out <- cfa(mod,HolzingerSwineford1939)
```

@@

```{r, message=FALSE, warning=FALSE,results='hide'}
library(regsem)
extractMatrices(out)["A"]
out.reg <- cv_regsem(out, type="lasso", 
                    pars_pen=c(1:9),n.lambda=15,jump=.05)
```

@@

```{r}
round(out.reg[[2]],3)
```

In this, the user can examine the penalty (lambda), whether the model converged ("conv"), with values of 0 indicating convergence, and the fit of each model. By default, two fit indices are output, both the root mean square error of approximation [RMSEA; @steiger1980], and the Bayesian information criteria [BIC; @schwarz1978estimating]. Both the RMSEA and BIC take into account the degrees of freedom of the model, an important point for model selection in the presence of lasso penalties (and other penalties that set parameters to zero). This concurs with work done on degrees of freedom for lasso regression, as @Zou2007 proved that the number of nonzero coefficients is an unbiased estimate of the degrees of freedom for regression. As the penalty increases, select parameter are set to zero, thus the degrees of freedom increases, which for fit indices that include the degrees of freedom in the calculation, means that although the likelihood of the model may only get worse (increase), both the RMSEA and BIC can improve (decrease) due to the impact of increased degrees of freedom.

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Ross Jacobucci\\University of Notre Dame
}
\title{\pkg{regsem}: Regularized Structural Equation Modeling}
\Keywords{regularization, structural equation modeling, latent variables, \proglang{R}}

\Abstract{
The \pkg{regsem} package in \proglang{R}, an implementation of
Regularized Structural Equation Modeling (Jacobucci, Grimm, and McArdle
2016), was recently developed with the goal incorporating various forms
of penalized likelihood estimation with a broad array of structural
equations models. The forms of regularization include both the
\textit{ridge} (Hoerl and Kennard 1970) and the least absolute shrinkage
and selection operator (\textit{lasso}; (Tibshirani 1996), along with
sparser extensions. The paper details both the algorithmic details and
an overview of the use of \pkg{regsem} through both a factor analysis
and latent growth curve models.
}

\Plainauthor{Ross Jacobucci}
\Plaintitle{regsem: Regularized Structural Equation Modeling}
\Shorttitle{\pkg{regsem}}
\Plainkeywords{regularization, structural equation modeling, latent variables, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
\Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Ross Jacobucci\\
  University of Notre Dame\\
  First line Second line\\
  E-mail: \href{mailto:rcjacobuc@gmail.com}{\nolinkurl{rcjacobuc@gmail.com}}\\
  URL: rjacobucci.com\\~\\
  }

\usepackage{amsmath} \usepackage{float} \usepackage{algorithm}
\usepackage[noend]{algpseudocode} \usepackage[subnum]{cases}

\begin{document}

\setlength\parindent{24pt} \setlength{\parskip}{0.5em}

\section{Introduction}\label{introduction}

The desire for simplicity in model structure comes by many names,
including simple structure (Thurstone 1935), variable complexity (Browne
2001), parsimony (Raykov and Marcoulides 1999; Marsh and Hau 1996),
``sparse loadings'' in the context of principal components analysis
(Zou, Hastie, and Tibshirani 2006), and lastly, ``sparsistency'',
denoting that all parameters in a sparse model that are zero are
correctly estimated as zero with probability tending to one (Lam and Fan
2009). The goal is accurately and efficiently estimate a model that is
parsimonious in allowing users to easily interpret the model's
representation of reality. In Figure 1, this would entail testing
whether the cross-loadings are necessary to the fit of the model, or
whether they can be pared away without an meaningful increase in bias.

@@

@@

As an example, a researcher may want to test this model, but may only
have a relatively small sample size (e.g.~80). However, there are 29
estimated parameters in this model, resulting in a estimated parameter
to sample size ratio as far below even the most liberal recommendations
(e.g.~10:1 parameters to sample size; Kline 2015). In lieu of finding
additional respondents, reducing the number of parameters estimated is
one effective strategy. Specifically, the 20 estimated regressions from
\textit{c1-c10} could be reduced to a number that makes the number of
parameters estimated to sample size ratio more reasonable.

\section{Regularization}\label{regularization}

Although a whole host of methods exist to perform variable selection,
the use of regularization has seen a wide array of application in the
context of regression, and more recently, in areas such as graphical
modeling, as well as a host of others. The two most common procedures
for regularization in regression are the \textit{ridge} (Hoerl and
Kennard 1970) and the least absolute shrinkage and selection operator
(\textit{lasso}; Tibshirani 1996); however, there are various
alternative forms that can be seen as subsets or generalizations of
these two procedures. Given an outcome vector \textit{y} and predictor
matrix \(X \in {R}^{n \times p}\) , the ridge estimates are defined as

\[\tag{1}
\hat{\beta}^{ridge}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p} \beta_{j}^{2}\Big\},
\] \noindent
where \(\beta_{0}\) is the intercept, \(\beta_{j}\) is the coefficient
for \(x_{j}\), and \(\lambda\) is the penalty that controls the amount
of shrinkage. Note that when \(\lambda = 0\), equation 3 reduces to
ordinary least squares regression. As \(\lambda\) is increased, the
\(\beta\) parameters are shrunken towards zero. The lasso estimates are
defined as

\[\tag{2}
\hat{\beta}^{lasso}= argmin \Big\{ \sum_{i=1}^{N} (y_{i} = \beta_{0} - \sum_{j=1}^{p}x_{ij} \beta_{j})^{2}  + \lambda \sum_{j=1}^{p}|\beta_{j}|\Big\}.
\]

\noindent
In lasso regression, the \(l_{1}\)-norm is used, instead of
\(l_{2}\)-norm as in ridge, which also shrinks the \(\beta\) parameters,
but additionally drives the parameters all the way to zero, thus
performing a form of subset selection.

In the context of our example depicted in Figure 1, to use lasso
regression to select among the covariates, the growth model would need
to be reduced to two factor scores, which neglects both the relationship
between both the slope and intercept, reducing both to independent
factor scores. Particularly in models with a greater number of latent
variables, this becomes increasing problematic. A method that keeps the
model structure, while allowing for a penalized estimation of specific
parameters is regularized structural equation modeling (RegSEM;
Jacobucci, Grimm, and McArdle 2016). RegSEM adds a penalty function to
the traditional maximum likelihood cost function for SEMs. The maximum
likelihood cost function for SEMs can be written as

\[\tag{3}
F_{ML}=log(\left|\Sigma\right|)+tr(C*\Sigma^{-1})-log(\left|C\right|)- p.
\]

\noindent
where \(\Sigma\) is the model implied covariance matrix, \(C\) is the
observed covariance matrix, and \(p\) is the number of estimated
parameters. RegSem builds in an additional element to penalize certain
model parameters yielding

\[\tag{4}
F_{regsem} = F_{ML} + \lambda P(\cdot)
\]

\noindent
where \(\lambda\) is the regularization parameter and takes on a value
between zero and infinity. When \(\lambda\) is zero, MLE is performed,
and when \(\lambda\) is infinity, all penalized parameters are shrunk to
zero. \(P(\cdot)\) is a general function for summing the values of one
or more of the model's parameter matrices. The two common forms of
\(P(\cdot)\) include both the Lasso (\(\| \cdot \|_{1}\)), which
penalizes the sum of the absolute values of the parameters, and Ridge
(\(\| \cdot \|_{2}\)), which penalizes the sum of the squared values of
the parameters.

In our example, the twenty regression parameters from the covariates to
both the intercept and slope would be penalized. Using lasso penalties,
the absolute value of these twenty parameters would be summed and after
being multiplied by the penalty \(\lambda\), added to equation 4,
resulting in:

\[\tag{5}
F_{lasso} = F_{ML} + \lambda * \left\|  \begin{matrix}  
c1 \xrightarrow[]{} i\\
c2\xrightarrow[]{}i\\
\vdots \\
c10\xrightarrow[]{}i\\
c1\xrightarrow[]{}s\\
\vdots \\
c10\xrightarrow[]{}s\\
\end{matrix}  \right\|_{1}
\]

@@

\subsection{Optimization}\label{optimization}

One method that has become popular for optimizing penalized likelihood
method is that of proximal gradient descent (e.g.~p.~104 in Hastie,
Tibshirani, and Wainwright 2015). In comparison to one-step procedures
common in SEM optimization, that only involve a method for calculating
the step size and the direction (typically using the gradient and an
approximation of the Hessian), proximal gradient descent can be
formulated as a two-step procedure. With a stepsize of \(\alpha^{t}\)
and parameters \(\theta^{t}\) at iteration \textit{t}:

@@

where \(S_{\alpha \lambda}(z)\) is the soft-thresholding operator
(Donoho 1995) used to overcome non-differentiability of the lasso
penalty at the origin:

\begin{equation}
S_{\alpha \lambda}(z_{j}) = sign(z_{j})(|z_{j}|-\alpha \lambda)_{+}
\end{equation}

where \((x)_{+}\) is shorthand for max(x,0) and \(\alpha\) is the step
size. Henceforth, \(\lambda\) is assumed to encompass both the penalty
and the step size \(s^{t}\). This procedure is only used to updated
parameters that are subject to penalty. Non-penalized parameters are
updated only using step 1 from above.

@@

\noindent

\begin{minipage}{\textwidth}
    %   \renewcommand\footnoterule{}    
    \begin{algorithm}[H]
        \begin{algorithmic}[1]
            %   \Procedure{RegSEM Block Coordinate Descent for Lasso}{}
            %   \footnotetext{Note: $A(pen)$ refers only to the penalized parameters, $vec()$ concatenates both vectors}
            \State Generate starting values for $\theta_{t}$
            \State Calculate initial fit $F_{t}$
            \State Set step size $\alpha$. .5 works well at this time.
            \State Set tolerance (tol). e.g. 1e-6
            \While {$|F_{t} - F_{t+1}| > tol$}
            \State Calculate gradient for A:$ \nabla (A) =:  \frac{\partial A}{\partial \theta_{t}}$
            \State $ \theta_{t+1^{*},A} =: \theta_{t,A} - \alpha \nabla(A)$
            \State  Update penalized parameters: $\theta_{t+1^{*},A(pen)} =: S_{\alpha \lambda}(\theta_{t+1^{*},A(pen)})$
            \State Calculate gradient for S:$ \nabla (S) =:  \frac{\partial S}{\partial \theta_{t+1^{*}}}$ 
            \State Update S parameters: $ \theta_{t+1^{*},S} =: \theta_{t,S} - \alpha \nabla(S)$
            \State $\theta_{t+1} =: \theta_{t+1^{*}}$ %vec(\theta_{t+1^{*},A},\theta_{t+1^{*},S})$
            \State Update $S_{t+1}, A_{t+1}$
            \State $ \Sigma_{t+1} = F(I-A_{t+1})^{-1}S_{t+1}(I-A_{t+1})^{-T}F^{T}$
            \State $ F_{t+1} = F_{ML}(\Sigma_{t+1},C) + \lambda \| A_{t+1}(pen) \|$
            \EndWhile
            %   \EndProcedure
        \end{algorithmic}
        \caption{RegSEM Block Coordinate Descent}
        \label{alg:seq}
    \end{algorithm}
\end{minipage}

\noindent
where step 13 is the calculation of the implied covariance matrix using
RAM matrices and step 14 calculates the fit of the model with penalizing
parameters in the \textit{A} matrix. Note that parameters from the
\textit{S} matrix can also be penalized, however, this is much less
common. This algorithm can be described as first order proximal block
coordinate descent. For some SEM models, using block updates has been
found to work better than standard gradient descent with the same soft
thresholding of penalized parameters.

\section{Types of Penalties}\label{types-of-penalties}

@@

\subsection{Elastic Net}\label{elastic-net}

Most notably, the elastic net (Zou and Hastie 2005) encompasses both the
ridge and lasso, reaching a compromise between both through the addition
of an additional parameter \(\alpha\).

\[
P_{enet}(\theta_{j}) = 0.5(1-\alpha)\| \theta_{j} \|_{2} + \alpha\| \theta_{j} \|_{2}
\]

with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
0,&  |\theta_{j}| < \alpha\lambda\\
\frac{sgn(\theta_{j})(|\theta_{j}|-\alpha\lambda)}{1+(1-\alpha)\lambda},              & |\theta_{j}|\geq\alpha\lambda
\end{cases}
\]

\noindent
When \(\alpha\) is zero, ridge is performed, and conversely when
\(\alpha\) is 1, lasso regularization is performed. This method
harnesses the benefits of both methods, particularly when variable
selection is warranted (lasso), but there may be collinearity between
the variables (ridge).

\subsection{Adaptive Lasso}\label{adaptive-lasso}

In using lasso penalties, difficulties emerge when the scale of
variables differ dramatically. By only using one value of \(\lambda\),
this can add appreciable bias to the resulting estimates (e.g. Fan and
Li 2001). One method proposed for overcoming this limitation is the
adaptive lasso (Zou 2006). Instead of penalizing parameters directly,
each parameter is scaled by the un-penalized estimated (maximum
likelihood estimate in SEM). This manifests itself as: \[
F_{alasso} = F_{ML} + \lambda \| \theta_{ML}^{-1} * \theta_{pen} \|_{1}
\] \noindent
with, following the same form for the lasso, the soft-thresholding
update is:
\href{High\%20dimensional\%20statistics\%20p.\%2032}{\textbackslash{}}:
\url{ftp://ftp.stat.math.ethz.ch/Teaching/buhlmann/advanced-comput-statist/notes1.pdf}
\[
S(\theta_{j})= sign(\theta_{j})(|\theta_{j}|-\frac{\lambda}{2|\theta_{j}|})_{+}
\] \noindent
In this, larger penalties are given for non-significant (smaller)
parameters, limiting the bias in estimating larger, significant,
parameters. Note that one limitation of this approach for SEM models is
that the model needs to be estimable with maximum likelihood.
Particularly for models with large numbers of variables, in relation to
sample size, this may not be possible.

\subsection{Smoothly Clipped Absolute Deviation
Penalty}\label{smoothly-clipped-absolute-deviation-penalty}

@@

Two additional penalties that overcome some of the deficiencies of the
lasso, producing sparser solutions, include the smoothly clipped
absolute deviation penalty (SCAD; Fan and Li 2001) and the minimax
concave penalty (MCP; Zhang 2010). In comparison to the lasso, both the
SCAD and MCP have much steeper penalties for smaller parameters, as is
evident in Figure **.

@@

\[
pen_{\lambda,\gamma}(\theta_{j}) = \lambda \big\{I(\theta_{j}\leq\lambda) + \frac{(\gamma \lambda-0)_{+}}{(\gamma-1)\lambda}I(\theta_{j}>\lambda)\big\}
\]

\noindent
with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
S(\theta_{j},\lambda),&  |\theta_{j}| \geq 2\lambda\\
\frac{\gamma-1}{\gamma-2}S(\theta_{j},\frac{\lambda\gamma}{\gamma-1}),              & 2\lambda < |\theta_{j}|\leq\alpha\lambda\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

@@

\subsection{Minimax Concave Penalty}\label{minimax-concave-penalty}

Zhang (2010) proposed the minimax concave penalty:

\[
pen_{\lambda,\gamma}(\theta_{j}) = \lambda\bigg(|\theta_{j}|-\frac{\theta_{j}^{2}}{2\lambda\gamma}\bigg)I(|\theta_{j}|<\lambda\gamma) +\frac{\lambda^{2}\gamma}{2}I(|\theta_{j}|\geq \lambda\gamma)
\]

\noindent
with a soft-thresholding update of \[
S(\theta_{j})= 
\begin{cases}
\frac{\gamma}{\gamma-1}S(\theta_{j},\lambda),&  |\theta_{j}| \leq \lambda\gamma\\
\theta_{j} & |\theta_{j}| > \lambda \gamma
\end{cases}
\]

@@

\section{Implementation}\label{implementation}

RegSEM is implemented as the \pkg{regsem} package in the \proglang{R}
statistical environment. To estimate the maximum likelihood fit of the
model, \pkg{regsem} uses \textit{Reticular Action Model} (RAM; J. J.
McArdle and McDonald 1984, McArdle (2005)) notation to derive an implied
covariance matrix. The parameters of each SEM are translated into three
matrices: the \textit{filter} (\textit{F}), the \textit{asymmetric}
(\textit{A}; directed paths; e.g.~factor loadings or regressions), and
the \textit{symmetric} (\textit{S}; undirected paths; e.g.~covariances
or variances). See Jacobucci, Grimm, and McArdle (2016) for more detail
on RAM notation.

Syntax for using the \pkg{regsem} is based on the \pkg{lavaan} package
(Rosseel 2012) for structural equation models. \pkg{lavaan} is a general
SEM software program that can fit a wide array of models with various
estimation methods. To use \pkg{regsem}, the user has to first fit the
model in \pkg{lavaan}. As an example, below is the code for a one factor
model fit in lavaan with the Holzinger-Swineford (Holzinger and
Swineford 1939) dataset.

\begin{CodeChunk}
\begin{CodeInput}
library(lavaan)
mod <- "
f1 = ~ NA*x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
f1~~1*f1
"
out <- cfa(mod,HolzingerSwineford1939)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
library(regsem)
extractMatrices(out)["A"]
out.reg <- cv_regsem(out, type="lasso", 
                    pars_pen=c(1:9),n.lambda=15,jump=.05)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
round(out.reg[[2]],3)
\end{CodeInput}
\begin{CodeOutput}
      lambda conv rmsea      BIC
 [1,]   0.00    0 0.188 7805.177
 [2,]   0.05    0 0.191 7815.389
 [3,]   0.10    0 0.199 7840.590
 [4,]   0.15    0 0.210 7875.842
 [5,]   0.20    0 0.203 7886.147
 [6,]   0.25    0 0.210 7911.816
 [7,]   0.30    0 0.214 7934.454
 [8,]   0.35    0 0.220 7960.670
 [9,]   0.40    0 0.227 7991.033
[10,]   0.45    0 0.229 8009.416
[11,]   0.50    0 0.234 8034.805
[12,]   0.55    0 0.242 8069.487
[13,]   0.60    0 0.286 8360.539
[14,]   0.65    0 0.286 8360.530
[15,]   0.70    0 0.286 8360.525
\end{CodeOutput}
\end{CodeChunk}
